Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-22 ResponseLetter.tex

Two places where you didn't do that, but that could really push this paper over the top into perfection territory, are the following: 1) offering your reader an explanation, even if brief, of the multi-type branching method central to your analyses/models, and why it was the right selection for this endeavor among other methods or evolutionary modeling frameworks used in the past or available to you,

\textcolor{blue}{ % change 53 
We have added a comment in line XXX: "We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population. Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection. In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress."
}

2) putting something real-world up against the parameter ranges you used to start the models, or just a brief statement about how they were established from the literature you drew upon (i.e., from experimental tests, patient data, other models, etc.). The latter is just in an effort to put something real to the numerical ranges that a reader can relate to before launching into the modeling.

\textcolor{blue}{ % change 54
We have updated the "Parameterization" section: "To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on Rew and Wilson (2000) and Bozic et al. (2013) for the division and death rates, respectively. 
Rew and Wilson (2000)  report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for sensitive and mutant cells. This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs. 
Bozic et al. (2013) report the growth rate $\Delta_s$ for sensitive melanoma cancer cells, from patient data, from which they deduce the death rate $0.11 \le \mu_s \le 0.17$. We use  $\mu_s=0.14$ per day. Additionally, they observed the growth rate of cancer cells before treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug. Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells." \\
We also have another paragraph in the same section about parameterization of the model using evidence from breast cancer.
}

@@

2. Do a search for multi-type branching and make sure you consistently use the hyphen or not.

\textcolor{blue}{%change 31
We have made the change to use the hyphen in all mentions of "multi-type branching" such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).
}

3. You identify that at least in certain scenarios aneuploidy plays an important role in the evolutionary rescue of tumor cell populations following chemotherapy, at least with one type of drug, and maybe limited to small primary or secondary tumors. But, we never really know the extent that this scenario is applicable. Is it specific to this particular chemotherapy, small tumors, only certain types of cancer? In conjunction with this, there are places in the manuscript where you say things like "for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue. Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care. If the answer is "I don't know how applicable this finding is or what the broader implications might be," you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically. It's ok to not know for sure, but you shouldn't pass the opportunity to talk about it in limits and future research.

\textcolor{blue}{
Thank you for this remark, you are correct to note that this statement is in opposition to some of our results. We have removed this statement and instead, we now write "The “stationary” scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting. Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig. 4A, inset), it could be biologically significant for some tumors under specific drug therapies. The current literature may be biased towards aneuploidies that provide drug resistance. 
Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
We also added the following sentences to the Discussion.
In line XXX: "Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$. \emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}" \\
And in line XXX: "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature. We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies."\\
The following sentences were already in the Discussion (line XXX): "In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation. 
Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
Together, this increases the cumulative number of mutants that arise." and "We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance."
}

### 2025-01-22 ms.tex
respond to last open comment

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{c}{\abs{\Delta_s}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is very large or the effect of the drug $\Delta_s$ is small. In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$. However, if the aneuploid growth rate, $\Delta_a$, increases, evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $\tilde{u}=10^{-3}$, and $v=10^{-7}$, we are in the first scenario (tolerant aneuploids) and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation. We note that for larger values of the pre-treatment chromosome missegregation rate, $\tilde{u}$, which are consistent with empirical studies (\Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutation.

@@

Using multi-type branching processes, we derived the probability of evolutionary rescue of the tumor under the effects of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.

@@

Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape. 
Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumor under specific drug therapies.

@@

\begin{figure}
\vspace*{1\baselineskip}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ProliferationTimeCDFN.pdf}
\end{subfigure}
\caption{\textbf{Aneuploidy extends the window of opportunity for evolutionary rescue.}
(A) The probability that a successful mutant has not appeared by time $t$. Black line: tolerant aneuploidy ($u>0, \lambda_a=0.0899$). Red line: stationary aneuploidy ($u>0, \lambda_a=0.089999$). Green line: partially resistant aneuploidy ($u>0, \lambda_a=0.095$). Blue line: no aneuploidy ($u=0$). Aneuploidy plays an important role in rescuing the tumor cell population as the sensitive population becomes extinct. Markers represent simulation results, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated, and $n=100$ is the number of simulations). Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$. %change 56
(B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$. Green line: $N=10^6$ (small tumor). Red line: $N=10^7$ (intermediate-sized tumor). Blue line: $N=10^{10}$ (large tumor). Increasing the initial tumor size guarantees that the cancer will relapse. Markers represent simulations, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
\label{cdffig}
\end{figure}

>>>>

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{c}{\abs{\Delta_s}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1\text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is very large or the effect of the drug $\Delta_s$ is small. In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$. However, if the aneuploid growth rate, $\Delta_a$, increases, evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $\tilde{u}=10^{-3}$, and $v=10^{-7}$, we are in the first scenario (tolerant aneuploids) and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation. We note that for larger values of the pre-treatment chromosome missegregation rate, $\tilde{u}$, which are consistent with empirical studies (\Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutation.

@@

Using multi-type branching processes, we derived the probability of evolutionary rescue of the tumor under the effects of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$. The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued (\Cref{figureAneuploidy}).

@@

Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape. 
Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of the drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
The current literature may be biased towards aneuploidies that provide drug resistance. 
Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (\Cref{rescue_prob}) and extend the window of opportunity for rescue (\Cref{cdffig}).

@@

\begin{figure}
\vspace*{1\baselineskip}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ProliferationTimeCDFN.pdf}
\end{subfigure}
\caption{\textbf{Aneuploidy extends the window of opportunity for evolutionary rescue.}
(A) The probability that a successful mutant has not appeared by time $t$. Black line: tolerant aneuploidy ($u>0, \lambda_a=0.0899$). Red line: stationary aneuploidy ($u>0, \lambda_a=0.089999$). Green line: partially resistant aneuploidy ($u>0, \lambda_a=0.095$). Blue line: no aneuploidy ($u=0$). Aneuploidy plays an important role in rescuing the tumor cell population as the sensitive population becomes extinct. Markers represent simulation results, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated, and $n=100$ is the number of simulations). Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$. %change 56
(B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$ when aneuploidy provides tolerance. Green line: $N=10^6$ (small tumor). Red line: $N=10^7$ (intermediate-sized tumor). Blue line: $N=10^{10}$ (large tumor). Increasing the initial tumor size guarantees that the cancer will relapse. Markers represent simulations, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
\label{cdffig}
\end{figure}
